---
title: "class09: Spooky Days"
author: "Caliope Marin (PID: A13912583)"
format: pdf
---

Today is 

```{r}
candy_file <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names=1)

head(candy_file)


```
>q.1 How many different candy types are in this dataset?

```{r}
nrow(candy_file)
```

#There are 85 candy types in this dataset

```{r}
table(candy_file["fruity"])
sum(candy_file["fruity"])
```

>Q2. there are 38 candies that are fruity

```{r}
candy_file["Reeses", ]$winpercent
```
>Q3  What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy_file["Air Heads", ] $winpercent
```


```{r}
library(dplyr)
candy_file |>
  filter(rownames(candy_file)%in% c("Dum Dums", "Twix")) |>
  select(winpercent)
```

>Q4. What is the winpercent value for “Kit Kat”?
>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy_file |>
  filter(rownames(candy_file)%in% c("Kit Kat", "Tootsie Roll Snack Bars")) |>
  select(winpercent)
#The win percent for Kit Kat is 76.76% and Tootsie Roll is # 49.65%
```

The %in% operator is useful for checking the intersection of two vectors

```{r}
c("barry","liz", "chandra") %in% c("paul", "alice", "liz")
```

```{r}
candy_file |>
  filter(winpercent > 75) |>
  filter(pricepercent < 0.5)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library(skimr)
skim(candy_file)
# Q6. The column that is on a different scale is the #winpercent because it is in percentage scale form
```

>Q7. What do you think a zero and one represent for the candy$chocolate column?
The "0" and "1" are binary for "False" and "true", respectively. 

>Q8. Plot a histogram of winpercent values

We can do this a few different ways: with base R `hist()`or with `ggplot(`)
```{r}
hist(candy_file$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy_file)+ 
  aes(winpercent)+
  geom_histogram(bins=10)+ 
  theme_bw()
```
>Q9. Is the distribution of winpercent values symmetrical?
>Q10. Is the center of the distribution above or below 50%?

#Q.9 no the distribution is not symmetrical 
#Q.10 the center of the distribution below 50%

```{r}
summary(candy_file$winpercent)
```

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?



```{r}
inds <-as.logical(candy_file$chocolate)
candy_file[inds,]$winpercent
```

```{r}
candy_file$chocolate==1
candy_file[inds,]$winpercent
```
```{r}
candy_file |>
  filter(chocolate==1)|>
  select(winpercent)
```
```{r}

candy_file |>
  filter(fruity==1)|>
  select(winpercent)
```
#Q.11 Yes chocolate has a higher preference than fruity candies

```{r}

inds<-candy_file$chocolate==1
choc.win <-candy_file[inds,]$winpercent

inds<-candy_file$fruity==1
fruit.win <-candy_file[inds,]$winpercent
#Then I could compare these
summary(choc.win)
summary(fruit.win)

```

```{r}
t.test(choc.win, fruit.win)
```


>Q12. Is this difference statistically significant?

#Yes the pvalue is really low -> 2.871-08 and chocolate
#is more preferred than fruity candy. #



```{r}
head(candy_file[order(candy_file$winpercent),], n=5)

```

There are two related functions that are useful here `sort()` and `order()`
```{r}
play <- c(2,1,5,3)
sort(play)
     
     
     
     
```

```{r}
order(play)
#Its giving you the number in the row of each variable not the actual character 
```

```{r}
play[order(play)]
```

```{r}
l<-c("c","a","b")
sort(l)

```


```{r}
order(l)
```

```{r}
n <-c("d","a")
n[order(n)]
```
>Q13. What are the five least liked candy types in this set?

```{r}
inds <-order(candy_file$winpercent)
head(candy_file[inds,])
#these are the least fave candies below
```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
#inds means the bucket of a vector
inds <- order(candy_file$winpercent, decreasing= T)
head(candy_file[inds,],5)
#these are the fave candies above
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy_file) + 
  aes(y=rownames(candy_file), x=winpercent ) +
  geom_col()
  

```
>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}

ggplot(candy_file) + 
  aes(winpercent, reorder(rownames(candy_file),winpercent)) +
  geom_col()
```

```{r}

#my fave candy is `Reese's Peanut Butter cup`
my_cols=rep("black", nrow(candy_file))

my_cols[as.logical(candy_file$chocolate)] = "chocolate"
my_cols[as.logical(candy_file$bar)] = "brown"
my_cols[as.logical(candy_file$fruity)] = "pink"
my_cols [rownames(candy_file)=="Reese's Peanut Butter cup"] <- "cyan"
```
>Q. Color your favorite candy

```{r}


ggplot(candy_file) + 
  aes(x=winpercent, y=reorder(rownames(candy_file),winpercent)) +
  geom_col(fill=my_cols) 
```


>Q17. What is the worst ranked chocolate candy?
#Sixlets 
>Q18. What is the best ranked fruity candy?
#Starbursts 

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy_file) +
  aes(winpercent, pricepercent, label=rownames(candy_file)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

#Reese's miniature's is the biggest bag for your buck

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
#Nik L Nip is the most expensive and least popular

```{r}
ord <- order(candy_file$pricepercent, decreasing = TRUE)
head( candy_file[ord,c(11,12)], n=5 )


```

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy_file) +
  aes(pricepercent, reorder(rownames(candy_file), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy_file), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
library(corrplot)
```


```{r}
cij <- cor(candy_file)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
#The anti-correlated variables are fruity and chocolate
>Q23. Similarly, what two variables are most positively correlated?
The highly correlated variables are chocolate and bar.

##Principal Component Analysis##

```{r}
pca <- prcomp(candy_file, scale=TRUE)
summary(pca)
```


```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```


```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy_file, pca$x[,1:3])

```


```{r}
barplot(pca$rotation[,1])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```



```{r}
#library(plotly)
#ggplotly(p)

```



```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")

```


>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
The variables are fruity, hard, and pluribus because those are anticorrelated with preference. 